High geomagnetic field intensity recorded by anorthosite xenoliths requires a strongly powered late Mesoproterozoic geodynamo

### 2022-02-12 PNAS_manuscript/PNAS_manuscript.tex

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$82 ZAm$^2$ for the virtual dipole moment from the anorthosites with the highest site-level values up to $\sim$127 ZAm$^2$ are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, our new results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core. The multiple observed paleointensity transitions from weak to strong in the Paleozoic and the Proterozoic present challenges on identifying the onset of inner core nucleation based on paleointensity records alone. 
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo regime at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was soon followed by initial crystallization of the inner core \cite{Bono2019a}. However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when we known the inner core exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously settle the timing of the initiation of inner core nucleation based on the current few paleointensity records.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are therefore an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma \cite{Swanson-Hysell2019a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks forms a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains and alteration during experiments is sagging or double-slopes as visualized in plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target a unique rock type---high-purity anorthosite xenoliths. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration with alteration of the plagioclase crystals themselves not resulting in the formation of secondary iron oxides in contrast with mafic minerals such as olivine. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal intrusions of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase at a paleolatitude of 22\textdegree $\pm$2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a, Zhang2021b}). Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals, and they do not display strong preferred orientations. These single-slope, high quality paleointensity data, together with other high-quality paleointensity records during the protracted Midcontinent Rift magmatic activity, require that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Magnetic field maps acquired with a quantum diamond microscope (QDM) on samples to which an IRM has been applied show that the relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite in (E) while large oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the overlapping magnetic fields above the oxides in (F). (E)-(H) show an experiment performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The isothermal remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that dipole-like remanent magnetizations of Beaver River anorthosite align well with the first applied field direction (E) and rotated to align with the second applied field direction with little anisotropic behavior (G). In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 tend to acquire an oblique remanence direction with respect to the field direction (F). After the application of a 300 mT external field, magnetization of those needles in plagioclase 1 did not reverse due to strong shape anisotropy whereas those needles in plagioclase 2 reversed but still acquired remanence oblique to the field direction (H). This experiment illustrates that the Beaver River anorthosite xenolith has much lower degrees of magnetic anisotropy than the Duluth anorthosite xenolith. B$_z$ denotes magnetic field intensity in the vertical direction with positive values corresponding to the out of page direction. Note the large contrast in magnetic field scale of (E) and (G) with those in (F) and (G). The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of large oxide inclusions with high aspect ratios (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers often exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G).

The features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these difference, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D, F, H). Magnetic imaging also finds that orientations of magnetic needles within the Duluth plagioclase crystals depict the orientation of remanence magnetizations as is shown by the oblique acquired remanence to the direction of a first applied isothermal remanent magnetization (IRM) in Figure \ref{fig:Petro_QDM}F. Upon applying a second IRM with a field of 300 mT in the orthogonal direction, the remanent magnetic moments failed to rotate in plagioclase 1, reversed in plagioclase 2 but still along the easy axis depicted by the needle orientations that are oblique to the applied field direction (Fig. \ref{fig:Petro_QDM}F, H). In contrast, the remanent magnetizations imaged in the Beaver River anorthosite xenoliths align much more parallel to the applied field directions upon applying a set of orthogonal external fields (Fig. \ref{fig:Petro_QDM}E, G). To our knowledge, this is the first time that the magnetic anisotropy associated with plagioclase petrofabrics is captured by magnetic imaging at such scale. This set of orthogonal IRM experiment clearly shows that the elongated (titano)magnetite exsolutions within plagioclase feldspar are responsible for the magnetic anisotropy of plagioclase which have been observed before at bulk sample scale \cite{Selkin2000a, Feinberg2006a}.

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} suggest that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single sites. Anorthosite xenoliths have paleointensity results that are consistent within small regions, but vary between regions. Anorthosite AX12 and AX16 in the Silver Bay area were emplaced $\sim$450 meters apart and have indistinguishable paleointensity estimates (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$10 km to the north, anorthosite xenoliths AX11 and AX13 in the Kennedy Creek area were also emplaced closely (separated by $\sim$125 meters) and yield similar values to one another but distinct paleointensity estimates from those of AX12 and AX16 (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$30 km further to the north in the Carlton Peak region, AX6, AX8, and AX10 are anorthosite xenoliths within 10 meters and also yield similar paleointensity estimates albeit with relatively large uncertainties (Fig. \ref{fig:PINT_cooling_corrected}). Therefore, one interpretation could consider the anorthosites of the three distinct regions to have captured at least three different records of geomagnetic field intensities during the emplacement and cooling of the Beaver River diabase sills ca. 1092 Ma. To contextualize our high paleointensity estimates, we plot the regional mean values based on specimen-level paleointensity results together with with the calculated paleointensity values at 22\textdegree N latitude based on the VADM values of model PADM2M \cite{Ziegler2011a} in Figure \ref{fig:PINT_cooling_corrected}B. The locality-mean values of these three regions are plotted in Figure \ref{fig:PINT_cooling_corrected}C where we also present the distribution of means calculated from 3 random values from the PADM2M model for a total of 10000 iterations. In addition, we perform the same comparison with the existing paleointensity data for the Cenozoic Era (the last 66 million years) from the PINT database (filtered by Q$_{PI}\geq$3) given that the observational dataset could record more geomagnetic field excursions or regional high flux patches (Fig. S5). The anorthosite xenoliths 3 region mean intensity is higher than all values from the resample results of the PADM2M model, and rival the highest values of the resample averages from the Cenozoic data (Fig. S5). These data supports that the anorthosite xenoliths likely record an exceptionally strong geomagnetic field ca. 1092 Ma.

>>>>

\begin{abstract}
Obtaining estimates of Earth's magnetic field strength in deep time is complicated by non-ideal rock magnetic behavior in many igneous rocks. In this study, we target anorthosite xenoliths that cooled and acquired their magnetization within ca. 1092 Ma shallowly emplaced diabase intrusions of the North American Midcontinent Rift. In contrast to the diabase which fails to provide reliable paleointensity estimates, the anorthosite xenoliths are unusually high-fidelity recorders yielding high-quality, single-slope paleointensity results that are consistent at specimen and site levels. An average value of $\sim$82 ZAm$^2$ for the virtual dipole moment from the anorthosite xenoliths, with the highest site-level values up to $\sim$127 ZAm$^2$, are higher than that of the dipole component of Earth's magnetic field today and rival the highest values in the paleointensity database. Such high intensities recorded by the anorthosite xenoliths require the existence of a strongly powered geodynamo at the time. Together with previous paleointensity data from other Midcontinent Rift rocks, these results indicate that a dynamo with strong power sources persisted for more than 14 million years ca. 1.1 Ga. These data are inconsistent with there being a progressive monotonic decay of Earth's dynamo strength through the Proterozoic Eon and could challenge the hypothesis of a young inner core. The multiple observed paleointensity transitions from weak to strong in the Paleozoic and the Proterozoic present challenges in identifying the onset of inner core nucleation based on paleointensity records alone. 
\end{abstract}

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to provide constraints on the thermal evolution of Earth’s core. Evidence for a persistent magnetic field through the Proterozoic, for example, likely necessitates the existence of plate tectonics that sustained core-mantle boundary heat flow \cite{Swanson-Hysell2021c}. However, strikingly low estimates of geomagnetic field strengths have been obtained ca. 565 Ma during the Ediacaran Period \cite{Bono2019a, Shcherbakova2019a,Thallner2021b} and ca. 370 Ma during the Devonian Period \cite{Shcherbakova2017a, Shcherbakova2021a, Hawkins2021a}, potentially indicating unusual periods of core dynamo activity at those times. The Ediacaran data have been interpreted to indicate that there was a progressively decaying field up to that time that was soon followed by initial crystallization of the inner core \cite{Bono2019a}. Sparse data in the Proterozoic Era (2500 to 539 Ma) and the reality of high variability in Phanerozoic Era records (539 to 0 Ma) necessitate additional data to evaluate interpreted trends.

% However, the ability for the geomagnetic dipole field to reach similar low values in rather recent times when we known the inner core exists (Fig. \ref{fig:PINT_compilation}) makes it challenging to unambiguously settle the timing of the initiation of inner core nucleation based on the current few paleointensity records.

@@

Determinations of the absolute value of ancient geomagnetic field strength rely on igneous rocks that acquire thermal remanent magnetizations as they cool. These magnetizations need to be unmodified by subsequent heating or chemical alteration in order to maintain a record of the ancient geomagnetic field from the time of cooling. Intracontinental magmatic events are an important target for determination of ancient paleointensity as they can be well-preserved within continental interiors. This interior position results in them typically being distant from tectonic events along continental margins that can drive alteration through heat and fluids. However, intraplate magmatism associated with large igneous provinces is typically of geologically short duration with the bulk of magmatic products emplaced within 1 Myr or less \cite{Kasbohm2021a}. The Midcontinent Rift (Fig. \ref{fig:Geologic_map}) is an exception as it is a large igneous province where magmatism lasted $\sim$25 Myr from \texit{ca.} 1109 Ma to 1084 Ma during which there were pulsed intervals of more rapid activity \cite{Swanson-Hysell2021a}. Additionally, extension ceased in the Midcontinent Rift prior to lithospheric separation, preserving volcanic, intrusive, and sedimentary rocks of the rift within the continental interior. As a result, rocks of the rift have unusually simple paleomagnetic behavior for their greater than one billion-year-old age and paleomagnetic data from rift rocks forms a central record of Mesoproterozoic paleogeography \cite{Swanson-Hysell2021c}. The duration of magmatic activity within the Midcontinent Rift is longer than the entire 20.4 Myr long Neogene Period such that it enables an extended well-preserved window into the intensity of Earth's magnetic field in the late Mesoproterozoic.

Despite the excellent preservation of the rocks, non-ideal paleointensity behaviors have challenged the interpretation of many previous paleointensity results from the Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a}. The most trusted type of paleointensity estimate is that obtained through experiments in which the primary natural thermal remanent magnetization is progressively replaced by a laboratory magnetization that is imparted in a known field with internal consistency checks (such as in IZZI-style Thellier experiments;  \citealp{Yu2004a}). In such Thellier paleointensity experiments, one typical departure from ideal behavior due to the presence of nonuniformly magnetized grains (with either multidomain \cite{Dunlop2001a} or vortex states \cite{Tauxe2020a}) is sagging or double-slopes as visualized in Arai plots that show thermal remanent magnetization (TRM) acquired versus natural remanent magnetization (NRM) lost. For such data, distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope. Typically, such non-ideal behavior would result in a higher paleointensity estimate from the steeper-sloped low-temperature portion and a lower paleointensity estimate from the high-temperature portion. For example, in data from the Midcontinent Rift, ref. \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Such non-ideal results were rejected by \citealp{Sprain2018a} who applied stricter paleointensity selection criteria, but as a result had few accepted sites.

In this study, we target a unique rock type---high-purity anorthosite xenoliths. Anorthosites are cumulate rocks composed dominantly of the mineral plagioclase. They are attractive targets for paleomagnetic study as plagioclase crystals can enshroud magnetic inclusions from alteration with alteration of the plagioclase crystals themselves not resulting in the formation of secondary iron oxides in contrast with mafic minerals such as olivine. The anorthosite xenoliths targeted in this study were brought to the near surface in magma that formed hypabyssal (shallowly emplaced) intrusions of the Beaver River diabase \cite{Zhang2021b}. They cooled and acquired their magnetizations in conjunction with the host diabase at a paleolatitude of 22\textdegree $\pm$2\textdegree (calculated from the paleomagnetic pole of the coeval Portage Lake Volcanics; \citealp{Swanson-Hysell2019a, Zhang2021b}). Paleointensity experiments on the anorthosite xenoliths have a high success rate, yielding consistent specimen- and site-level paleointensity results. Rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly within the range of relevant field strengths. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers within and interstitial to plagioclase crystals, and they do not display strong preferred orientations. These single-slope, high-quality paleointensity data, together with other high-quality paleointensity records during the protracted Midcontinent Rift magmatic activity, require that there was a strong late Mesoproterozoic geodynamo.

@@

While magmatic activity within the Midcontinent Rift was protracted, there were intervals of particularly rapid volcanism and voluminous emplacement of intrusions \cite{Swanson-Hysell2021a}. The ca. 1092 Ma Beaver Bay Complex in northern Minnesota punctuates one such period of magmatism during the main stage of Midcontinent Rift activity. The magma that formed the 1091.7$\pm$0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. These anorthosite xenoliths are plagioclase cumulates that formed comagmatically with the host diabase in the lower crust---an interpretation confirmed by U-Pb zircon geochronology \cite{Zhang2021b}.

@@

%\begin{figure*}
\begin{SCfigure*}
\centering
\noindent\includegraphics[width=13 cm]{Petro_QDM_full.png}
\caption{\footnotesize{Thin section petrographic images (A,B,C,D) and magnetic maps (E,F) of anorthosite from the Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (left column;  \citealp{Zhang2021b}) and a distinct anorthosite from the Duluth Complex Anorthositic Series (right column). The Duluth Complex anorthosites were not targeted for paleointensity experiments in this study given the complexities associated with more pronounced fabrics. Cross-polarized petrographic images of the Beaver River anorthosite (A,C) reveal plagioclase with a granoblastic texture with crystals that are largely free of large opaque inclusions. In contrast, plagioclase crystals in the Duluth Complex anorthosite have euhedral, interlocking crystals with an igneous foliation (B) and the plagioclase crystals contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the [001] axis of the plagioclase. Maps of the vertical component of magnetic field (B$_z$) developed with a quantum diamond microscope (QDM) show relatively weak magnetic sources within and interstitial to plagioclase crystals in the Beaver River anorthosite (in E) relative to the strongly magnetic large oxide needles within Duluth Complex plagioclase (in F). The B$_z$ color scale is an order of magnitude greater in the Duluth Complex anorthosite maps (F, H) than the maps for the Beaver River anorthosite (E, G). (E to G) and (F to H) show experiments performed on both samples where we apply a first field along the y axis of the field of view and then apply a second field of 300 mT orthogonal to the first field direction. The remanent magnetizations acquired by both anorthosites were mapped after the application of each field. The magnetic images show that remanent magnetizations of Beaver River anorthosite (i.e. the individual dipoles visible with paired red +B$_z$ and blue -B$_z$ lobes) align well with the first applied field direction (E) and then rotated to align with the second applied field direction indicating minimal anisotropic behavior (G). In contrast, the remanence magnetizations carried by the magnetic needles in plagioclase 1 of the Duluth anorthosite xenolith align well with the first applied field but those in plagioclase 2 acquired an oblique remanence direction with respect to the field direction (F). After the application of an orthogonal 300 mT external field, magnetization of those needles in plagioclase 1 did not change direction due to strong shape anisotropy whereas the magnetization of needles in plagioclase 2 flipped, but with the acquired remanence remaining oblique to the field direction (H). This experiment illustrates that the Beaver River anorthosite xenolith has much lower magnetic anisotropy than the Duluth anorthosite xenolith. The unit of the axes in the QDM maps are in $\mu$m.}}
\label{fig:Petro_QDM}
%\end{figure*}
\end{SCfigure*}

@@

The dominantly monomineralic anorthosite xenoliths within the Beaver River diabase often have granoblastic texture characterized by equigranular crystals with weakly developed petrofabrics (Fig. \ref{fig:Petro_QDM}A). In addition, the plagioclase crystals of the anorthosite xenoliths are typically free of oxide inclusions 10s of microns or greater in length. with high aspect ratios (Fig. \ref{fig:Petro_QDM}C). Magnetic imaging using a quantum diamond microscope (QDM) shows that the dominant magnetic remanence carriers exist within, and sometimes interstitial to, plagioclase crystals (Fig. \ref{fig:Petro_QDM}E, G).

The features of the Beaver River anorthosite xenoliths are distinct from many other anorthosites such as the anorthosite xenoliths within the Duluth Complex Anorthositic Series rocks---older intrusions within the Midcontinent Rift. To illustrate these differences, we present petrographic and magnetic imaging data from a sample of Duluth Complex anorthosite that was not targeted for paleointensity experiments. The plagioclase of the Duluth Complex anorthosites typically develop interlocking textures that display pronounced igneous foliation (Fig. \ref{fig:Petro_QDM}B). In addition, there are abundant Fe-Ti oxide needles that are typically aligned with the [001] axes of the plagioclase crystals (Fig. \ref{fig:Petro_QDM}D). Magnetic imaging confirms that these strongly magnetic needles within the plagioclase crystals often have magnetic moments oriented along their long axes (Fig. \ref{fig:Petro_QDM}D, F, H). Magnetic imaging also finds that orientations of magnetite needles within the Duluth plagioclase crystals dictate the orientation of remanence magnetizations as is shown by magnetization being acquired at angles highly oblique to applied isothermal remanent magnetization (IRM) directions (Figure \ref{fig:Petro_QDM}F). In an experiment where a  second orthogonal IRM is applied, the remanent magnetic moments either fail to rotate the magnetization of the ferromagnetic inclusions in a given plagioclase grain or flip the magnetization by 180º remaining in a direction that is oblique to the applied field direction dictated by shape anisotropy (Fig. \ref{fig:Petro_QDM}F, H). These experiments enable novel visualization of the grain-scale magnetic anisotropy of elongated exsolved (titano)magnetite inclusions within plagioclase that leads to magnetic anisotropy observed at the bulk sample scale \cite{Selkin2000a, Feinberg2006a}. In contrast, the remanent magnetizations of ferromagnetic grains imaged in the Beaver River anorthosite xenoliths, targeted for paleointensity in this study, align with the applied field directions and rotate in the experiment where an orthogonal field is applied (Fig. \ref{fig:Petro_QDM}E, G). These results highlight that the population

%To our knowledge, this is the first time that the magnetic anisotropy associated with plagioclase petrofabrics is captured by magnetic imaging at such scale.

%Those new IRM maps look really spectacular- extremely convincing that your anorthites are very different and much less anisotropic than the usual ones!  I would just add some text to the paper (and maybe the supplement if you run out of space) explaining these data and saying that this really shows what anisotropy looks like on the grain scale, i.e. dipoles oriented towards the easy axis most aligned to the IRM direction.  I don't think I've seen such clear grain-scale B field maps of extreme anisotropy published elsewhere, so it would be good to emphasize this is a really novel measurement and technique for seeing the origin of anisotropy.

@@

\subsection*{Considering secular variation and cooling rate}
To best characterize the geomagnetic axial dipole field intensity during a certain time period a paleointensity dataset should cover a sufficient amount of time such that paleosecular variations of the geomagnetic field are averaged. Thermal modeling results from ref. \citealp{Zhang2021b} suggest that the Beaver River anorthosite xenoliths were heated to tholeiitic magma temperature ($\sim$1100\textdegree C) and acquired thermal remanent magnetization during cooling with their diabase host on a time scale of a thousand years, partially averaging secular variation within single sites. Anorthosite xenoliths have paleointensity results that are consistent within small regions, but vary between regions. Anorthosite AX12 and AX16 in the Silver Bay area were emplaced $\sim$450 meters apart and have indistinguishable paleointensity estimates (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$10 km to the north, anorthosite xenoliths AX11 and AX13 in the Kennedy Creek area were also emplaced closely (separated by $\sim$125 meters) and yield similar values to one another but distinct paleointensity estimates from those of AX12 and AX16 (Figs. \ref{fig:Geologic_map} and \ref{fig:PINT_cooling_corrected}). $\sim$30 km further to the north in the Carlton Peak region, AX6, AX8, and AX10 are anorthosite xenoliths within 10 meters and also yield similar paleointensity estimates albeit with relatively large uncertainties (Fig. \ref{fig:PINT_cooling_corrected}). Therefore, one interpretation could consider the anorthosites of the three distinct regions to have captured at least three different records of geomagnetic field intensities during the emplacement and cooling of the Beaver River diabase sills ca. 1092 Ma. To contextualize our high paleointensity estimates, we plot the regional mean values (Table. S3) based on specimen-level paleointensity results together with with the calculated paleointensity values at 22\textdegree N latitude based on the VADM values of model PADM2M \cite{Ziegler2011a} in Figure \ref{fig:PINT_cooling_corrected}B. The locality-mean values of these three regions are plotted in Figure \ref{fig:PINT_cooling_corrected}C (and reported in Table. S3) where we also present the distribution of means calculated from 3 random values from the PADM2M model for a total of 10000 iterations. In addition, we perform the same comparison with the existing paleointensity data for the Cenozoic Era (the last 66 million years) from the PINT database (filtered by Q$_{PI}\geq$3; Fig. S5) given that the observational dataset could record more geomagnetic field excursions or regional high flux patches. The anorthosite xenoliths 3 region mean intensity is higher than all values from the resample results of the PADM2M model, and rival the highest values of the resample averages from the Cenozoic data (Fig. S5). These data supports that the anorthosite xenoliths likely record an exceptionally strong geomagnetic field ca. 1092 Ma.
